% !TEX root = ../bachelor-thesis.tex

\chapter{Conclusions}
\label{chap:conclusion}

The core objective of this research was to investigate whether a minimal, low-cost sensor suite -- comprising a 2D LiDAR, GPS, IMU, and wheel encoders -- combined with semantic environmental data and a CPU-optimized trajectory planner is sufficient for safe, autonomous UGV navigation in semi-structured outdoor environments. By balancing the need for capable outdoor autonomy with the limitations of constrained hardware, this thesis successfully demonstrated that heavy sensors and RAM-intensive SLAM algorithms are not strict prerequisites for reliable point-to-point navigation.

The global path planning objective was fulfilled by constructing a semantic OpenStreetMap environment graph and applying the deterministically optimal A* algorithm over it. By filtering OSM access tags and surface materials, collapsing multi-segment physical paths into navigable edges, and projecting geodetic coordinates into a local Cartesian frame, the system successfully generated metric-accurate global routes. This entirely eliminated the need to build or maintain a spatial map at runtime, validating the use of existing semantic data for lightweight global routing.

The obstacle detection objective was fulfilled through a custom 2D LiDAR perception pipeline. Raw polar scans were filtered and clustered based on physical clearance thresholds, abstracting the environment into compact Cartesian geometric primitives -- lines and circles -- using Split-and-Merge and an adapted circumcenter fitting technique. Empirically, this abstraction constrained the system's memory footprint to an average of $293\text{ MB}$ of RAM, standing in stark contrast to the $4-16\text{ GB}$ typically demanded by dense visual SLAM pipelines~\cite{schmidt2025vislam}.

The local navigation objective was accomplished by integrating a dual-EKF localization setup with a custom Python-based Timed Elastic Band trajectory optimizer. The PyCeres solver was optimized for real-time execution through analytical Jacobians, temporal warm-starting, and explicit latency compensation, operating under the constraints of single-threaded execution. In a highly dynamic, pedestrian-dense simulation stress test, the local planner achieved an $83.3\%$ success rate while maintaining a $9.99\text{ Hz}$ median control frequency and demanding less than $13\%$ peak CPU utilization on a consumer-grade mobile processor, demonstrating that kinodynamically feasible MPC-style optimization is achievable within strict hardware bounds.

Taken together, these results demonstrate that robust obstacle avoidance and trajectory optimization can be achieved without a dedicated GPU or heavy 3D sensors, enabling the deployment of accessible, energy-efficient UGV platforms. Practically, this lightweight architecture is well suited for operations such as automated park maintenance or campus delivery in semi-structured pedestrian zones. The system's safety architecture -- evidenced by the successfully recovered dynamic contact events during testing -- further demonstrates that a hierarchical software-level emergency stop approach provides a practically sound safety net even when the primary planner's prediction horizon is breached.

The failure mode analysis, however, outlines the current architectural limitations and informs directions for further research. The deadlock failures confirm that a single-topology elastic band planner struggles in shrinking corridors, and the logical next step is the integration of an extended TEB formulation that simultaneously maintains and optimizes multiple candidate trajectories~\cite{rossman2017teb}, allowing the solver to select an alternative route around uncooperative agents rather than becoming trapped in a single corridor. The rear-collision failure underscores a critical vulnerability of unidirectional 2D perception: safe bidirectional motion requires either 360-degree sensor coverage or enforcement of a strict forward-only kinematic constraint. Finally, while the $83.3\%$ success rate is promising, the 95\% Wilson score interval ($66\%-93\%$) remains broad due to the sample size of 30 runs. Future work should encompass larger-scale quantitative trials and physical outdoor benchmarking to tighten these bounds and explicitly measure EKF dead-reckoning drift under severe GPS occlusion, such as that induced by dense tree canopies.
