% !TEX root = ../bachelor-thesis.tex

\chapter{Experiments and Results}
\label{chap:experiments}

\section{Experimental Objective}
\label{sec:experimental-objective}

The goal of the experimental evaluation was to assess the proposed lightweight navigation pipeline under dynamic obstacle conditions. As detailed in Section~\ref{sec:experimental-methodology}, the primary benchmarking was conducted in simulation using a $50$-meter stress test course populated with perpendicularly moving pedestrians. This setup was chosen to explicitly test the limits of the TEB local planner's elastic deformation capabilities and the reliability of the system's safety nets. The evaluation focused on three primary aspects: navigation success, computational efficiency, and hardware-software integration.

\section{Navigation Results}
\label{sec:navigation-results}

A total of 30 independent simulated runs were executed on the generated dynamic course. The UGV successfully completed the course in 25 runs, yielding a success rate of 83.33\%. Table~\ref{tab:navigation-outcomes} details the distribution of outcomes, including specific failure modes and recoverable safety interventions.

\begin{table}[htpb]
    \centering
    \caption{Summary of Navigation Outcomes}
    \label{tab:navigation-outcomes}
    \begin{tabular}{lcc}
        \hline
        \textbf{Outcome} & \textbf{Count} & \textbf{Percentage} \\
        \hline
        \textbf{Successful runs} & \textbf{25} & \textbf{83.33\%} \\
        \quad With recoverable safety events & 9 & 30.00\% \\
        \textbf{Failed runs} & \textbf{5} & \textbf{16.67\%} \\
        \quad Squeeze/deadlock failures & 4 & 13.33\% \\
        \quad Rear blind-spot collision failure & 1 & 3.33\% \\
        \hline
    \end{tabular}
\end{table}

Out of the 25 successful runs, 9 runs included at least one recoverable safety event. In these instances, a moving pedestrian entered the critical stop radius of the robot, triggering the proximity breach safety net. The robot halted all motion according to its safety constraints, the simulated pedestrian approached and touched part of the robot, but once the pedestrian passed, the vehicle resumed its optimized trajectory. These events were not categorized as failures because the safety hierarchy functioned exactly as intended, preventing robot-caused collisions, and the robot ultimately completed its navigational objective.

\subsection{Travel Time and Path Efficiency}
\label{subsec:results-travel-time}

Motion metrics were calculated exclusively across the 25 successful runs, as shown in Table~\ref{tab:motion-results}. To provide a comprehensive representation of system performance without the distortion of extreme outliers, the distribution of results is reported using means, standard deviations, quartiles (Q1, Q3), and the 5th and 95th percentiles (P5, P95).

\begin{table}[htpb]
    \centering
    \caption{Travel Time and Path Efficiency for Successful Runs}
    \label{tab:motion-results}
    \begin{tabular}{lccccccc}
        \hline
        \textbf{Metric} & \textbf{Mean} & \textbf{Std.} & \textbf{P5} & \textbf{Q1} & \textbf{Median} & \textbf{Q3} & \textbf{P95} \\
        \hline
        Travel Time (s) & 145.73 & 33.54 & 110.60 & 128.80 & 140.40 & 150.10 & 225.30 \\
        Path Efficiency & 0.658 & 0.089 & 0.490 & 0.616 & 0.656 & 0.717 & 0.788 \\
        \hline
    \end{tabular}
\end{table}

The mean travel time was $145.73\text{ s}$, with a standard deviation of $33.54\text{ s}$. This high variance strongly reflects the dynamic nature of the environment: runs where pedestrians crossed unfavorably required the UGV to execute extensive detours or pause for several seconds, whereas favorable timing allowed for near-continuous forward progress. 

The mean path efficiency was $0.658$. Because this metric is the ratio of the shortest A* path to the actual executed distance, the result indicates that the robot traveled approximately $52\%$ further ($1 / 0.658 = 1.52$) than the ideal route. This extra distance is entirely expected, as the local planner was forced into heavy lateral maneuvering to avoid the dense pedestrian traffic.

\section{Computational Performance}
\label{sec:computational-performance}

A primary objective of this thesis was to demonstrate that autonomous navigation can be achieved with low resource overhead. Telemetry was logged continuously during the simulation to profile the Python-based controller execution. The results are summarized in Table~\ref{tab:resource-results}.

\begin{table}[htpb]
    \centering
    \caption{Computational Resource Usage and Control Frequency}
    \label{tab:resource-results}
    \begin{tabular}{lccccccc}
        \hline
        \textbf{Metric} & \textbf{Mean} & \textbf{Std.} & \textbf{P5} & \textbf{Q1} & \textbf{Median} & \textbf{Q3} & \textbf{P95} \\
        \hline
        Control Freq. (Hz) & 9.19 & 2.24 & 3.69 & 9.88 & 9.99 & 10.03 & 10.36 \\
        CPU Utilization (\%) & 9.26 & 2.43 & 5.00 & 7.50 & 9.49 & 11.63 & 11.99 \\
        Memory (MB) & 292.99 & 3.31 & 287.34 & 291.52 & 293.92 & 295.71 & 297.29 \\
        \hline
    \end{tabular}
\end{table}

The TEB controller was designed to target an update rate of $10\text{ Hz}$. The measured median frequency was $9.99\text{ Hz}$, with an interquartile range (Q1 to Q3) tightly bounded between $9.88\text{ Hz}$ and $10.03\text{ Hz}$, confirming that the system reliably met the intended cycle time. The 5th percentile value ($3.69\text{ Hz}$) indicates occasional optimization latency, corresponding to mathematically complex solver frames where multiple dynamic obstacles entered the prediction horizon simultaneously.

Resource consumption remained exceptionally light. Process CPU utilization averaged just $9.26\%$, peaking at $12.36\%$ on the test system. Memory footprint demonstrated extreme stability, averaging $292.99\text{ MB}$ with a maximum standard deviation of only $3.31\text{ MB}$, confirming that geometric obstacle abstraction effectively bypasses the heavy RAM requirements associated with dense costmaps or SLAM pipelines.

\section{Failure Mode Analysis}
\label{sec:failure-analysis}

The five failed runs provided critical insights into the inherent limitations of the proposed architecture. The failures were categorized into two distinct mechanisms: squeeze deadlocks and blind-spot collisions.

\subsection{Squeeze and Deadlock Failures}
Four of the five failures occurred when the UGV attempted to navigate between a pedestrian and the virtual boundary representing the curb. As the pedestrian moved laterally toward the curb, the physical free space progressively collapsed. The system's safety layer correctly identified the proximity breach and stopped the vehicle to prevent a collision. However, once stopped, the TEB local planner failed to recover, being too closely located to curb to risk resuming the navigation.

This behavior confirms a structural limitation of elastic-band-based planning algorithms. Because TEB optimizes by iteratively deforming a continuous trajectory matrix, it fundamentally struggles to abandon its current topological class. When trapped in a shrinking corridor, the optimizer cannot easily ``snap'' the trajectory to the opposite side of the pedestrian. The robot therefore remained frozen in a deadlock state, requiring manual operator intervention.

\subsection{Rear Blind-Spot Collision}
The final failure occurred during an attempted reverse maneuver. Faced with a similar squeeze scenario, the TEB optimizer generated a trajectory that commanded the robot to drive backward to escape the collapsing space. Unfortunately, a pedestrian happened to be located directly behind the vehicle. Because the Husky relies exclusively on a single 2D LiDAR with a 270-degree forward-facing field of view, the rear region was completely unobserved by the perception pipeline. The UGV therefore executed a kinodynamically valid, but physically disastrous, reversing trajectory into a blind spot, resulting in an unrecoverable collision. 

\section{Qualitative Hardware Validation}
\label{sec:hardware-validation}

While the primary quantitative benchmarking was conducted in simulation to guarantee reproducibility, a preliminary physical deployment was executed using the real Clearpath Husky A200 platform. The objective of this test was to qualitatively validate the hardware-software integration, specifically confirming that the perception pipeline processes real-world sensor noise and that the motor controllers correctly execute the generated TEB trajectories.

Because physical testing took place in an indoor laboratory setting, GPS and OSM global routing were unavailable. The test was therefore designed to evaluate the local trajectory optimization and sensor fusion in isolation. The EKF was configured to rely entirely on the IMU and wheel odometry for state estimation, while the local planner was provided with a relative $10\text{-meter}$ Cartesian goal directly ahead of the UGV. 

Four physical square obstacles, each approximately $1.5\text{ m}$ in length, were placed in a staggered pattern between the robot and the goal. The maximum linear velocity was constrained to a conservative $0.2\text{ m/s}$ for safety during the initial indoor deployment.

The UGV successfully navigated the course. The custom perception pipeline accurately clustered the raw Hokuyo 2D LiDAR data and abstracted the physical boxes into line and circle primitives. The TEB optimizer continuously generated kinodynamically feasible maneuvers, successfully steering the robot around the staggered obstacles and reaching the target destination without triggering the emergency stop safety nets. 

While a subsequent, unrelated technical issue with the Husky A200 platform prevented the execution of comprehensive outdoor quantitative testing, 
% TODO: mention the issue or not?
this physical run serves as a proof of concept. It confirms that the underlying ROS 2 architecture functions correctly on physical hardware, that the computational latency of the Python-based Ceres solver is acceptable for real-time control, and that the abstraction of real, noisy LiDAR data into geometric primitives is a viable approach for physical obstacle avoidance.

\section{Discussion}
\label{sec:discussion}

The experimental results validate the core hypothesis of the thesis: safe, autonomous point-to-point navigation in semi-structured environments is feasible utilizing only lightweight, geometric obstacle abstractions and minimal sensor inputs. The qualitative indoor hardware validation confirmed that the pipeline successfully translates real sensor data into physical motor commands. Furthermore, the quantitative simulation stress test demonstrated an 83.33\% success rate in a deliberately adverse dynamic environment, achieved using under $13\%$ CPU and $300\text{ MB}$ of RAM. This underscores the efficiency of bypassing dense visual SLAM and grid-based mapping. The robustness of the hierarchical safety nets was particularly evident, recovering from 9 distinct dynamic contact events.

However, the stress test also explicitly highlighted the boundary limits of this minimal approach. The squeeze failures demonstrate that relying purely on local elastic deformation is insufficient for complex crowd navigation. Without an explicit higher-level recovery behavior to force topological replanning or predict obstacle velocities, the UGV is vulnerable to deadlocks. Furthermore, the blind-spot failure clearly establishes that while 2D perception is computationally efficient, executing bidirectional trajectories mandates either 360-degree sensor coverage or strict forward-only motion constraints.
