% !TEX root = ../bachelor-thesis.tex

\chapter{Introduction}
\label{chap:introduction}

\section{Background and Motivation}
\label{sec:background-motivation}

Mobile robotics has become increasingly important for applications involving monotonous, remote, or hazardous tasks. Manual teleoperation of such systems is often infeasible due to scalability constraints, communication latency, and signal degradation in obstructed environments. Consequently, deploying autonomous systems capable of independent path finding and target navigation provides a more robust operational paradigm.

Outdoor navigation introduces significant complexities compared to indoor environments, as systems must account for less predictable conditions, including moving obstacles, uneven terrain, and non-cooperative dynamic agents.

Currently, off-the-shelf software frameworks for autonomous navigation exist, such as the Nav2 stack in ROS~\cite{nav2_docs}. However, modern outdoor systems often rely on resource-intensive sensors (like 3D LiDARs with data rates of millions of points per second, and RGB-D cameras) and complex algorithms (like Visual SLAM~\cite{schmidt2025vislam} and MPPI~\cite{williams2016mppi}).
% O.F. Оце heavy sensors мене напрягає, зізнаюся -- можуть виникнути питання. Можливо, варто тут більше написати подробиць, окреслити спектр рішень і де Ваше поміж них. Маса-точніст-енергоспоживання-потоки даних etc
% R.H. Виправив
These approaches may provide superior accuracy in unstructured environments but demand substantial computational resources, large amounts of RAM (often $4 -- 16$ GB), and sometimes GPUs. This increases costs, drains battery life, and is not always possible on simple, low-cost platforms.

Therefore, there is a strong motivation to investigate the capabilities of a minimal, lightweight sensor suite for navigation. Mitigating the reliance on computationally intensive perception pipelines enables more accessible, cost-effective, and energy-efficient robotic platforms.

Instead of generating a map from scratch using SLAM, existing semantic data such as OpenStreetMap (OSM) can be used for global routing~\cite{hentschel2010osm}. Instead of processing 3D point clouds or images, a simple 2D LiDAR can handle immediate, dynamic obstacles like pedestrians or trees. Finally, instead of using visual odometry, the system can rely on standard GPS fused with an IMU and wheel odometry for localization~\cite{moore2015ekf}.

Together, these lightweight components provide all the necessary data for a robot to navigate, albeit with lower spatial fidelity. This thesis aims to demonstrate that such a minimal setup allows for safe, real-time, collision-free point-to-point navigation in a semi-structured outdoor park environment, achieving high success rates while respecting strict CPU and memory bounds.
% O.F. Як на мене, тут варто конкретніше про те, що демонструється -- на додачу до сказаного
% R.H. Виправив

The setup includes a differential-drive ground robot in an outdoor park. Pedestrian paths provide a semi-structured, topologically known environment. Unlike road networks, pedestrian zones lack strict traffic rules or lane changes, significantly reducing the need for heavy semantic computer vision processing.
% O.F. Може то занадто сильно, що аж eliminating?
% R.H. Виправив
However, unlike completely unstructured off-road environments~\cite{min2026offroad}, these paths have clear geometric boundaries that can be modelled and tracked. The system's objective is to navigate autonomously from a start point to a target point without human intervention.

\section{Objectives}
\label{sec:objectives}

% O.F. Щось воно мені зовсім як "цілі" не звучить... Це ніби цілі робота, не Ваші цілі описано, а потім Ви зразу про результати. Чи потрібно детальніше пояснення, про що я?
% R.H. Виправив

The primary objective of this thesis is to design, implement, and evaluate a resource-efficient autonomous navigation system for a UGV in semi-structured outdoor environments. Specifically, the research addresses the following goals:

\begin{itemize}
\item Develop global path planning using the A* search algorithm on an OpenStreetMap environment graph.
\item Design obstacle detection by processing 2D LiDAR data to extract and classify geometric primitives (lines and circles).
\item Implement local navigation by combining spatial localization (via an Extended Kalman Filter fusing GPS, IMU, and wheel odometry) with Timed Elastic Band trajectory optimization to follow the global path while avoiding obstacles.
\end{itemize}

The system was validated in a simulated environment, with performance evaluated based on success rate, safety margins, and travel time.

\section{Structure Of The Thesis}
\label{sec:structure}

The remainder of this thesis is structured as follows:

\begin{itemize}
    \item \textbf{Chapter~\ref{chap:related-work} (Related Work)} surveys existing approaches to path planning, localization, semantic mapping, and obstacle abstraction, highlighting the gap in lightweight solutions for outdoor navigation.
    \item \textbf{Chapter~\ref{chap:theoretical-background} (Theoretical Background)} details the underlying concepts, including sensor characteristics, robot kinematics, hierarchical navigation, trajectory optimization, sensor fusion, and geometric feature extraction.
    \item \textbf{Chapter~\ref{chap:proposed-solution} (Proposed Solution)} describes the hardware setup, system architecture, software implementation (including OSM processing, LiDAR abstraction, EKF localization, and custom TEB optimizer), safety mechanisms, and ethical considerations.
    \item \textbf{Chapter~\ref{chap:experiments} (Experiments and Results)} evaluates the proposed architecture through dynamic simulation stress tests and qualitative hardware validation, analyzing navigation success rates, computational efficiency, and specific failure modes.
    \item \textbf{Chapter~\ref{chap:conclusion} (Conclusions)} summarizes the fulfillment of the research objectives, discusses the practical implications of the lightweight architecture, and outlines current system limitations alongside directions for future research.
\end{itemize}

